Joint maximum-likelihood of phylogeny and ancestral states is not consistent

### 2019-03-19 versions/response.v2.txt

> The language used to refer to the approach and references to the
> literature are awkward. The references to profile likelihood work
> suggest that there is a theoretical basis for the approach (that paper
> suggests consistency) but the settings considered in such work are
> quite different, I believe. Integrated likelihood usually refers to
> something different too. Finally, as Reviewer 3 makes clear the only
> current implementation of the approach was intended as an
> approximation to ML rather than as ML. The central issue is whether
> treating unobserved data (ancestral states here) as parameters
> sometimes works; the EM algorithm were developed with the idea that
> this is not what should be done. In any case, references directly tied
> to that issue would be more relevant. Additional suggestions include:
> 1) Make a clear statement in the intro about why the usual likelihood
> is considered the right likelihood: it is the probability of the
> *observed* data. 2) The point that large numbers of parameters need
> not defeat likelihood methods is valid but it should be made clear
> when referencing such work that the setting is different.

D TODO

> -l21: Some rewording is needed. The joint method will, as n gets
> large, continue to estimate a well-resolved topology with positive
> probability; albeit with small middle edge-length.

It will estimate a branch length of zero which is equivalent to a multifurcating topology.

@@

> 1) "jointly estimating phylogenetic parameters along with ancestral
> states is not consistent in general"... "only correct on a set of
> measure 0".
> It is clearly true that branch length will be biased. But the fact
> that joint estimation will not recover the true branch lengths is not
> surprising but perfectly expected. Joint estimation ignores subleading
> configurations of internal states. Since the subleading terms
> typically require more state changes, branch length estimates ignoring
> them are biased. In a similar way a Taylor expansion is almost nowhere
> "correct", but still a useful and insightful approximation (a better
> analogy is ignoring looped diagram in a diagrammatic expansion.).

E TODO

> 2) In some parameter regions the internal branch of the four taxon
> tree collapses to 0.
> Again, this is true and identifying regions in parameter space where
> the internal branch collapses to zero is the most interesting part of
> the manuscript in my opinion. Unfortunately, the manuscript doesn't
> give much insight as to why the length of the internal branch shrinks
> to zero in some regions of parameter space. There is, however, a
> rather intuitive explanation for this. In the inv Fels topology, the
> internal branch collapses to 0 if
>
> p_x^2 > c p_y
>
> (the prefactor c depends on specifics like alphabet size etc). And
> this has an intuitive interpretation: if p_x^2 > p_y, it is more
> likely to observe a mutation twice on two branches of length p_x than
> once on a branch of length p_y. As a result, the joint inference will
> push p_w to zero and assign all state changes on the terminal branches
> of length p_x. But for short branches, this corresponding sector of
> parameter space is a small (tending to zero for a fixed p_x/p_y with
> p_x \to 0).
>
> My empirical tests of the approximate joint inference scheme suggest
> this region is much smaller than the one identified by Shaw et al in
> Fig 3.

We have worked this out with Dr. Neher and found that the difference lay in that he was not optimizing the objective function from both sides.
This also demonstrated that the above-listed intuitive explanation was incorrect.

### 2019-03-20 versions/response.v2.txt

> The reviewers have universally viewed the revision as much improved.
> That said, 3/4 maintain that further changes are necessary prior to
> publication. It is our view that the remaining criticisms are "minor",
> and should you address them thoroughly and convincingly, we expect
> that final acceptance of the revision can be done editorially.

Thank you for handling the manuscript, and please thank the reviewers for their helpful comments. As stated by Reviewer 3, we believe that the review process has significantly improved the manuscript.

@@

> Reviewers' comments:
> Reviewer: 1

> Comments to the Author
> As with the previous version, I believe the topic matter will be of
> interest. For me, however, the work is incomplete in important ways
> that I believe can be relatively easily addressed. The authors went
> beyond my expectations in calculating explicit solutions for the
> limiting edge-lengths (Fig S2) when tau1 is the putative topology of
> estimation. This allows them to draw conclusions about edge-length
> inconsistency but there is no definitive conclusion about `strict'
> topological inconsistency (in the sense of converging upon the wrong
> tree rather than a multifurcating one). From the statements near l238
> it seems the authors have completed some preliminary investigation on
> strict topological inconsistency and find no evidence for it at
> present. Assuming that remains true, the findings of the current ms
> are of a different nature than the previous version where strict
> topological inconsistency was found. In settings with a limiting
> multifurcating topology, because it is multifurcating, for finite but
> large sequence-lengths, measures of uncertainty (eg. bootstrap
> support) will suggest that there is substantial uncertainty in
> whatever topology gets estimated (this probably deserves discussion).
> The joint optimization approach thus lacks the power to estimate the
> correct topology where ML would but is not being positively mislead.
> Assuming the lack of strict topological inconsistency holds, the
> results for the method are not as negative as they were in the
> previous version.

%EM I honestly can't figure out what the reviewer wants to see re the bootstrap. Branch length inferences don't come into the bootstrap unless you contract things with zero branch lengths, which isn't generally done when doing bootstrap estimation.

The reviewer is correct that this story is different than the one in the previous version of the paper.

@@

> -l154 and (should check elsewhere): `the limiting value of the
> estimator' rather than `the estimator' is what is being referred to
> here.

DISCUSS

> -l186: which `likelihood terms' is not that clear to me: `each site
> results in multiple factors, only one of which depends on w with
> dependence being through (1+w) or (1-w) depending ...'

We have clarified this by specifying that we're talking about the joint likelihood expression here.

@@

> -l189: `depends on the values of other branch parameters' or some
> other rewording. I don't think it is that they are unknown. It is that
> they need optimization, so you can't optimize w independently of them.

We have adopted the suggested wording, thanks.

> -l201, Figs 2 and 3: I don't understand why direct numerical
> optimization was needed. You have the exact expressions for hat w in
> Fig S2. So you could have plotted them as Fig 2. A similar comment
> applies to Fig 3. Are you trying to convince the reader that your
> proofs have no errors? That seems more an issue for supp material.

%EM David, am I right with this?
Although we have expressions for the likelihoods, it is not trivial to infer the branch lengths maximizing those likelihoods.

### 2019-03-20 joint_inf.tex
Start on responses and edits.

In words, the joint estimation procedure does not recover the true generating $t^*$ almost everywhere in the space of generating parameters.
The proof can be seen intuitively through the fact that the estimator for a given branch fidelity cannot isolate the individual generating parameter as a linear term since the estimator itself is a combination of nonlinear functions in $(x^*,y^*)$ of generating fidelities.
For example, in estimating $x_1$, $\hat{x}_1$ is a linear combination of $p_{\siteSplit_j}$ values.
For the generating probabilities (top panel of Table~\ref{tab:gen-sitepatprob}), there is no linear combination that results in an isolated $x^*$ term, as all terms are either quadratic, i.e., $(x^*)^2$ or $(x^*)^2(y^*)^2$, or have both $x^*$ and $y^*$, i.e., $x^*(y^*)^2$.
Thus, we cannot obtain a linear combination that in the general case yields $x^*$.
We may have special cases where certain $x^*$ and $y^*$ values yield consistent estimates for $\hat{t}$, but consistency does not hold in general.
We give an example of consistency in a degenerate case in the Appendix following the proof of Theorem~\ref{thm:bl_theorem}.

@@

This result implies an inconsistency as we estimate the interior branch length to be zero (i.e., interior branch fidelity is one) in an open set of values for $x^*$ and $y^*$ (Fig.~\ref{fig:bl-general-inconsistency}).
Although the regions seem coarse, $\hat{w}$ has a closed form in the entire space of $x^*$ and $y^*$ (table in bottom panel of Fig.~\ref{fig:max-anc-state}).
As we consider different topologies $\tau_1$ and $\tau_2$ for $\hat{t}$, the incorrect topology $\tau_2$ attains a likelihood value at its maximum equal to that of the true topology $\tau_1$ in the limit.
In other words, if $w=1$ the objective functions $\ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)$ and $\ell_{\tau^*,t^*}(\tau_2, t; \boldsymbol\xi)$ are equivalent.
We elaborate on this point in the Appendix.
The proof is through analytically reducing the general case to 81 separate cases (Table~\ref{tab:likelihoods}) to obtain a closed form maximal value for each.

We provide the following as an intuition for the theoretical development.
For a particular site pattern, to obtain the joint maximum likelihood function we maximize over ancestral states.
For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of our likelihood terms depending on which ancestral state corresponds to the highest conditional log-likelihood.
As $(1+w) > (1-w)$, a maximization procedure tends to prefer the $(1+w)$ term, though this is not guaranteed because the maximum depends on the values of the unknown branch parameters $t$.
Nevertheless, this tendency to include $(1+w)$ terms in the likelihood results in a positive bias of branch fidelities, i.e., estimating branch lengths to be shorter than truth.
This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which the most likely ancestral states are the same for each internal node letting $x_1=x_2=x^*$ and $y_1=y_2=y^*$ ($\xi_j=\emptyset$ for all $j$ in Table~\ref{tab:likelihoods}).
If we allow multifurcating trees in our inference, then we can think of this as an instance of converging to the wrong topology, as the true $y^*\neq 1$.

@@

We have shown an inconsistency when performing joint inference on branch lengths given an InvFels topology and investigated the performance of branch parameter estimation.
There is substantial scope for future work to make these results more precise and more general.
All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases, such as applied examples with larger trees or more realistic mutation models that are of interest to practitioners.
Also, given that many of the bounds presented here are in the form of level sets of multivariate polynomials, a more formal approach using algebraic geometric techniques may reveal more stable or interesting patterns of inconsistency; see \citet{Sturmfels2002} for a thorough treatment of solving systems of polynomial equations.
Finally, all of the material presented here concerns joint estimation under maximum likelihood, and does not pose any problem for other settings, such as joint sampling of trees and ancestral sequences in a Bayesian framework.

\section*{Acknowledgements}
We thank Richard Neher, Vladimir Minin, and Joe Felsenstein for helpful discussions.
Insight from the reviewers and the editors greatly improved this manuscript.

>>>>

In words, the joint estimation procedure does not recover the true generating $t^*$ almost everywhere in the space of generating parameters.
The proof that $\hat{t} \neq t^*$ everywhere except a set of measure zero can be seen intuitively through the fact that the estimator for a given branch fidelity cannot isolate the individual generating parameter as a linear term since the estimator itself is a combination of nonlinear functions in $(x^*,y^*)$ of generating fidelities.
For example, in estimating $x_1$, $\hat{x}_1$ is a linear combination of $p_{\siteSplit_j}$ values.
For the generating probabilities (top panel of Table~\ref{tab:gen-sitepatprob}), there is no linear combination that results in an isolated $x^*$ term, as all terms are either quadratic, i.e., $(x^*)^2$ or $(x^*)^2(y^*)^2$, or have both $x^*$ and $y^*$, i.e., $x^*(y^*)^2$.
Thus, we cannot obtain a linear combination that in the general case yields $x^*$.
We may have special cases where certain $x^*$ and $y^*$ values yield consistent estimates for $\hat{t}$, but consistency does not hold in general.
We give an example of consistency in a degenerate case in the Appendix following the proof of Theorem~\ref{thm:bl_theorem}.

@@

This result implies an inconsistency because the joint estimate of the interior branch length is zero (i.e., interior branch fidelity is one) in an open set of values for $x^*$ and $y^*$ (Fig.~\ref{fig:bl-general-inconsistency}).
$\hat{w}$ is available in closed form in the entire space of $x^*$ and $y^*$ (table in bottom panel of Fig.~\ref{fig:max-anc-state}).
As we consider different topologies $\tau_1$ and $\tau_2$ for $\hat{t}$, the incorrect topology $\tau_2$ attains a likelihood value at its maximum equal to that of the true topology $\tau_1$ in the limit.
In other words, if $w=1$ the objective functions $\ell_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi)$ and $\ell_{\tau^*,t^*}(\tau_2, t; \boldsymbol\xi)$ are equivalent.
We elaborate on this point in the Appendix.
The proof is through analytically reducing the general case to 81 separate cases (Table~\ref{tab:likelihoods}) to obtain a closed form maximal value for each.

We provide the following as an intuition for the theoretical development.
For a particular site pattern, to obtain the joint maximum likelihood function we maximize over ancestral states.
For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of the terms of the joint likelihood expression depending on which ancestral state corresponds to the highest conditional log-likelihood.
As $(1+w) > (1-w)$, a maximization procedure tends to prefer the $(1+w)$ term, though this is not guaranteed because the maximum depends on the values of the other branch parameters $t$.
Nevertheless, this tendency to include $(1+w)$ terms in the likelihood results in a positive bias of branch fidelities, i.e., estimating branch lengths to be shorter than truth.
This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which the most likely ancestral states are the same for each internal node letting $x_1=x_2=x^*$ and $y_1=y_2=y^*$ ($\xi_j=\emptyset$ for all $j$ in Table~\ref{tab:likelihoods}).
If we allow multifurcating trees in our inference, then we can think of this as an instance of converging to the wrong topology, as the true $y^*\neq 1$.

@@

We have shown an inconsistency when performing joint inference on branch lengths given an InvFels topology and investigated the performance of branch parameter estimation.
There is substantial scope for future work to make these results more precise and more general.
All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases, such as applied examples with larger trees or more realistic mutation models that are of interest to practitioners.
Initial results by of Dr\. Neher using the TreeTime \citep{Sagulenko2018-xl} package has shown that the region of parameter space on which joint inference infers a zero internal branch length for 4-state DNA models is smaller than that for 2-state binary models.
Also, given that many of the bounds presented here are in the form of level sets of multivariate polynomials, a more formal approach using algebraic geometric techniques may reveal more stable or interesting patterns of inconsistency; see \citet{Sturmfels2002} for a thorough treatment of solving systems of polynomial equations.
Finally, all of the material presented here concerns joint estimation under maximum likelihood, and does not pose any problem for other settings, such as joint sampling of trees and ancestral sequences in a Bayesian framework.

On a practical note, we would like to emphasize that TreeTime has implemented marginal mode inference since the publication of the \citet{Sagulenko2018-xl} paper.
In January 2019 a warning was added, in all caps, that is triggered when joint inference is used for trees such that any branch length is greater than 0.15.

\section*{Acknowledgements}
We thank Richard Neher, Vladimir Minin, and Joe Felsenstein for helpful discussions.
We are especially thankful to Dr.\ Neher for providing TreeTime simulation code that allowed us to test our conclusions on simulated data.
Insight from the reviewers and the editors greatly improved this manuscript.
